### 2014-01-23 rscripts/manuscript.Rnw
assorted updates

% PLOS One TEMPLATE
\usepackage{amsmath,amssymb,graphicx,cite,setspace}
\usepackage[textwidth=1.75in]{todonotes}
\usepackage[total={6in,9.5in},top=0.5in,left=0.5in,includefoot]{geometry}
\usepackage{listings}
\usepackage{url}
\usepackage{color}

@@

%\renewcommand{\lstlistingname}{Scala Code}

%\doublespacing

@@

\begin{document}
\SweaveOpts{concordance=TRUE}
\begin{flushleft}
{\Large
\textbf{Modeling Requirements for Simulating Covert Social Groups}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carl A. B. Pearson$^{1,\ast}$, 
Edo Airoldi$^{2}$,
Edward Kao$^{2}$,
Burton Singer$^{1}$, 
\\
\bf{1} Emerging Pathogens Institute, University of Florida, Gainesville, FL, USA
\\
\bf{2} Statistics, Harvard University, Cambridge, MA, USA
\\
$\ast$ E-mail: cap10@ufl.edu
\end{flushleft}
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  Moreover, increasing active observation would be expected to perturb these social systems.  However, these groups never have perfect cover, their secrecy may make them conspicuously absent from the everyday activity of the background population, and finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc. -- that might predict their actions.

This situation -- strong constraints on passively and actively obtained direct empirical data, overwhelming indirect data, and competing mechanical explanations -- recommends simulation.  However, because the of the censoring issues in the data, that simulation requires particular attention to characterizing uncertainty and avoiding overfitting.

Herein, we discuss these issues in terms of network models.  Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) finally some philosophy relative to implementation.

As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.  We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g. Receiver Operator Characteristic).  Finally, we also demonstrate a small extension.

@@

\section*{Introduction}
For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.  In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.  Of course that revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.  Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.

The underlying drive for these opposed efforts is the implacably expanding byte trail.  Once, the recorded information of an entire life might amount only to a few bytes -- parish records on births and deaths, perhaps including morbidity -- but now information era tools produce a near endless supply or 1s and 0s.  In some cases, this production is a side effect of organized commercial instinct -- monetizing search through targetted advertisement, e.g.  In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.  However, in many cases this bitwise production is an engineering consequence of the associated technology.  E.g., cellular phones transmit constant location data, financial systems are increasingly just data-based, and of course any use of the internet produces veritable bit contrails.  This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.

Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.  We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.  Like ``algorithmic'' bubble prediction, we tend to side more with Fama than Shiller.

Given the real uncertainty, what these filters call for is testing and validation, but those present their own difficulties.  Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' ranges from incomplete to deceptive, and experiments could have dangerous side effects.  Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without even further abstraction), even assuming the historical data are more than victor's embellishment.

Generating synthetic data seems like an obvious alternative.  It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.  Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.

@@

\subsection*{Modeling the Covert Groups}
Relative to the covert group, the simulation must model how the group members act and interact.  Typically, that further entails a modeling few moving pieces: (i) the members internal state (including what other members they know and communicate with), (ii) how that internal state evolves relative to external forces or states, and (iii) what actions those members undertake based on their internal state (possibly in response to some specific external event).

@@

An alternative representation for the background would be to model it as a continuous entity vice individuals in a network.  This could be used to reduce the simulation size.  However, this model would still need to create activity data that was consistent (in format) to that created by the covert group, because of the simulation requirements for observation.

@@

\section*{A Language Matching the Simulation Requirements: Scala}
Scala is an mixin inheritance object oriented, functional language, which is also includes an message-passing based model for parallel computing (which is transitioning from the initial implementation to that developed by the Akka project).  Indeed, Scala is used under the hood for the network simulation framework NetLogo.

@@

Second, many interactions in an organization can be thought of as management directing labor to complete a task; in a functional paradigm, functions are themselves treated as objects.  This translates natural to sending a function object from one simulated individual (in a leadership position) to another (in a subordinate position).  Loosely adopting some of the other functional paradigms, as Scala does -- emphasis on immutability, principally -- provides practical advantages to concurrent programming, which is necessary for large scale simulations.

Finally, the message-passing framework is an exact analogy to an individual / agent / actor - based model.  Having this baked into the language, makes it a feature more like control loops or collections, than a more complicated syntax available only advanced programmers (sometimes available on via library support) rather than more phenomena and model focused scientists.

@@

In considering the salafi network, we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to ``Good'' or ``Bad''.  The communication connections between individuals will be static, therefore we may set the social network at the outset of the simulation.

@@

\subsection*{The Covert Group}
First, we define the covert group context:

\begin{code}
  def plotter(
    superiors:Set[Plotter] = Set.empty,
    collaborators:Set[Plotter] = Set.empty,
    subordinates:Set[Plotter] = Set.empty) : Receive = {
      case _ => // do nothing with messages
  }
\end{code}

We need to define how they adopt and evolve this context.  Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though be taking this approach we have laid the ground for monitoring that activity as part of future detection scenarios.  The simulation runner will send a message \texttt{Radicalize} to the initial cabal, containing all its members:

\begin{code}
  case class Radicalize(id:Long, leaders:Set[Plotter])

  def receive = {
    case Radicalize(_, cabal) =>
      context become plotter(collaborators = (cabal - self))
    case _ => // do nothing with other messages
  }
\end{code}

@@

Some of these new recruits might be introduced to others as collaborators, so we need to need to handle collaboration messages, again assuming that the simulation runner is setting the particular covert structure.

\begin{code}
  case class Collaboration(id:Long, members:Set[Plotter])

  def plotter(...) = {
    case Collaboration(_, members) =>
      context become plotter(superiors, collaborators ++ members, subordinates)
    // ... other cases
  }
\end{code}

Similarly, we can advise these recruits about other actors for them to recruit.

Then we need to consider how these actors will know about each other in the simulation, how they will interact, etc.  In this example, we consider a static group, and as such the simulation could simply initialize the group.  However, we know that these groups do not really appear fully formed, and thus someday it will likely be desirable to include recruitment in the model.  We can anticipate that day by building our static initialization in those terms.

We will represent recruitment activity with two messages, which we add to the definition of the Leader:

\begin{code}[title=Covert Leader Helper Object]
object Leader {
  case class Recruit(id:Long) // try to recuit a member
  case class RecruitAck(id:Long) // join the group
}
\end{code}

We then modify the definitions of Leader and Member to respond to these messages:

\begin{code}[title=Using Messages]
class Leader extends Actor {
  var recruits = Set.empty[Member]

  def receive = {
    case RecruitAck(_) =>
      recruits += sender
    case _ =>
  }
}

class Member extends Actor {
  var leader;

  def receive = {
    case Recruit(id) =>
      leader = sender
    case _ =>
  }

}
\end{code}

@@

>>>>

\usepackage{amsmath,amssymb,graphicx,cite,setspace} % PLOS One TEMPLATE
\usepackage{listings, url, color, appendix} % things I usually use
% items to make todonotes useful
\usepackage[textwidth=1.75in]{todonotes} 
\usepackage[total={6in,9.5in},top=0.5in,left=0.5in,includefoot]{geometry}
\newcommand{\wantFig}[2]{
  \begin{figure}[h]\centering
    \missingfigure{}
    \caption{#2}
    \figlabel{#1}
  \end{figure}
}
% alt layout formatting
%\doublespacing

@@

@@

\newcommand{\figlabel}[1]{\label{fig_#1}}
\newcommand{\figref}[1]{fig. \ref{fig_#1}}
\newcommand{\Figref}[1]{Fig. \ref{fig_#1}}

\newcommand{\applabel}[1]{\label{app_#1}}
\newcommand{\appref}[1]{appendix \ref{app_#1}}
\newcommand{\Appref}[1]{Appendix \ref{fig_#1}}

\begin{document}
\SweaveOpts{concordance=TRUE}
\begin{flushleft}
{\Large
\textbf{Modeling Requirements for Simulating Covert Social Groups}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carl A. B. Pearson$^{1,\ast}$, 
Edo Airoldi$^{2}$,
Edward Kao$^{2}$,
Burton Singer$^{1}$, 
\\
\bf{1} Emerging Pathogens Institute, University of Florida, Gainesville, FL, USA
\\
\bf{2} Statistics, Harvard University, Cambridge, MA, USA
\\
$\ast$ E-mail: cap10 \{at\} ufl \{dot\} edu
\end{flushleft}
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  Moreover, increasing active observation would be expected to perturb these social systems.  However, these groups never have perfect cover.  Their secrecy may make them conspicuously absent from the everyday activity of the background population.  Finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc. -- that might predict their actions.

The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.  With any covert network, there are strong constraints on passively and actively obtainable direct empirical data, overwhelming amounts of indirect data, and competing mechanical explanations for what is observed.  Studying the detailed history of any one network, or set of them, will not necessarily provide insight about how to effectively detect such networks in the future.  This situation recommends simulation of a variety of scenarios with accompanying detection strategies. Because of the censoring issues in the data, simulation requires particular attention to characterizing uncertainty and avoiding overfitting.

Herein, we discuss these issues in terms of network models.  Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) some philosophy relative to implementation.

As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.  We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g. Receiver Operator Characteristic).  Finally, we also demonstrate a small model extensionand show how it affects the (unextended) detection performance.

@@

\section*{Introduction}
For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.  In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.  Of course the revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.  Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.

The underlying drive for these opposed efforts is the implacably expanding byte trail.  Once, the recorded information of an entire life might have amounted only to a few bytes -- e.g., parish records on births and deaths, perhaps including morbidity.  Now information era tools produce a near endless supply of 1s and 0s.  In some cases, this production is a side effect of organized commercial instinct -- Google monetizing search through targeted advertisement, for example.  In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.  However, in many cases this bitwise production is an engineering consequence of the associated technology.  E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers with almost no physical exchange, and of course any use of the internet produces veritable bit contrails.  This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.

Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.  We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.  Like ``algorithmic'' bubble prediction, we side more with Fama than Shiller.

Many such filters are based on social network descriptions.  These social networks can be a compact way to represent measured relations in a population.  However, these approaches all boil down to measuring properties of mathematical representation, with those properties having a tenuous connection to a mechanistic explanation of those phenomena.

Given the real uncertainty, what these filters call for is testing and validation, but those present their own difficulties.  Calling field testing ``problematic'' seems like a gross understatement; reference ``truth'' ranges from incomplete to deceptive, and experiments could have dangerous side effects.  Even making use of intensely studied historical events is problematic: these offer no way to consider evolutionary behavior and technological innovation (without an additional layer of abstraction that would create an even more complex empirical challenge), even assuming the historical data are more than victor's embellishment.

Generating synthetic data seems like an obvious alternative.  It allows for comparison across both detection and masking strategies, consideration of multiple background contexts, forecasting of risks and tradeoffs in a way that allows uncertainties, and in general providing a framework for imaginative assessment.  Like all such flexible tools producing quantitative results, it has the subtle downside of the simulator's biases being validated by numerical gospel; if one believes a particular strategy is effective -- perhaps even with reasonable agreement for a particular time and situation -- there would be a natural tendency to ``adjust'' scenarios until they indicated the success of strategy.  There have been other attempts at such an approach\cite{carley2003dynamic}.

@@

\subsection*{Modeling the Covert Groups}
Relative to the covert group, the simulation must model how the group members act and interact.  Typically, that further entails modeling a few moving pieces: (i) the members internal state (including what other members they know and communicate with), (ii) how that internal state evolves relative to external forces or states, and (iii) what actions those members undertake based on their internal state (possibly in response to some specific external event).

@@

An alternative representation for the background would be to model it as a continuous entity instead of individuals in a network.  This could be used to reduce the simulation size.  However, this model would still need to create activity data that was consistent (in format) to that created by the covert group, because of the simulation requirements for observation.

@@

\section*{A Language Matching the Simulation Requirements: Scala}
Scala\cite{odersky2004overview} is an mixin-inheritance\cite{bracha1990mixin}, object-oriented, functional language, which also includes an message-passing based model for parallel computing\footnote{The actor model is transitioning from the initial implementation to that developed by the Akka project\cite{akka}.  We base our discussion on the Akka model.  The most important difference is the notion of contexts and context switching: with context switching, an actor alters how it responds to events.  This is a useful abstraction for representing evolving individual state.}.  Indeed, Scala is used under the hood for the network simulation framework NetLogo.

@@

Second, many interactions in an organization can be thought of as management directing labor to complete a task; in a functional paradigm, functions are themselves treated as objects.  This translates natural to sending a function object from one simulated individual (in a leadership position) to another (in a subordinate position).  Loosely adopting some of the other functional paradigms, as Scala does -- emphasis on immutability, principally -- provides practical advantages to concurrent programming, which is necessary for large scale simulations.  Additionally, the functional emphasis bakes in other programming constructs to do with collection processing (e.g., iteration, transformation, filtering) that are less convenient in earlier languages.

Finally, the message-passing framework is an exact analogy to an individual / agent / actor - based model.  Having this baked into the language makes it a feature more like control branches (e.g., \texttt{if}, \texttt{else}, \texttt{for}) or collections, than a more complicated syntax available only to advanced programmers (and often only via library support) rather than more phenomena- and model-focused scientists.

@@

In considering the salafi network (see \figref{salafi_network}), we will observe a highly constrained set of activity: only communication, with all of that communication having been reduced to \texttt{Good} or \texttt{Bad}.  The communication connections between individuals will be static, therefore we may set the social network at the outset of the simulation.

\wantFig{salafi_network}{the salafi network}

@@

\subsection*{The Covert Group}
First, we define the covert group \texttt{Receive} context:

\begin{code}
  def plotter(
    // a plotter has superiors, collaborators, and subordinates
    // these groups are initially empty
    superiors:Set[Plotter] = Set.empty,
    collaborators:Set[Plotter] = Set.empty,
    subordinates:Set[Plotter] = Set.empty) : Receive = {
      // ... to-be-defined behavior
      case _ => // ignore unmatched messages
  }
\end{code}

We need to define how they adopt and evolve this context.  Since we are not studying the organizational dynamics with this simulation, we can be phenomenologically loose for this part, though this approach means we can add monitoring that activity as part of future detection scenarios with minimal coding.  The simulation runner will send a \texttt{Radicalize} message to the initial cabal, then a \texttt{Collaboration} message to the same group:

\begin{code}
  case class Radicalize(id:Long)
  case class Collaboration(id:Long, group:Set[Plotter])

  def receive = {
    case Radicalize(_) =>
      context become plotter()
    case _ => // ignore other messages
  }

  def plotter(...) => {
    case Collaboration(_, group) =>
      context become plotter(collaborators = collaborators ++ (group - self))
    // ...
  }
\end{code}

@@

At this point, we have all the necessary messages to construct the given network.  Those messages, based on the labels in Sageman et al. are covered in \Appref{sage_network}.  However, we still need to model how the plotters actually plot.  We will assume that all plotting (i) originates with leadership, (ii) when intermediates receive direction, they work with their \texttt{collaborators} as well as delegating work, (iii) that directions occasionally result in feedback to higher-ups, and finally (iv) all of the ``work'' and direction is captured as passing \texttt{Bad} messages.

@@

\begin{appendices}
\section{Messages to Construct Salafi Network}\applabel{sage_network}
\begin{code}
  // ...TODO figure out these messages
  val cabal = Set(``A'',``B'',``C'');
  cabal foreach { _ ! Radicalize(...) }
  cabal foreach { _ ! Collaboration(..., cabal) } 
\end{code}

\end{appendices}
